---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library("jsonlite")
library("ggplot2")
library(tidyr)
library(dplyr)
library(lubridate)
```

```{r}
stockfish = read.csv("lichess_stockfish_sample.csv")
# nofish = read.csv("lichess_nofish_sample.csv")
```

```{r}
summary(stockfish)
```

```{r}
game = stockfish[3, 'moves']

# Extract the eval and movetime columns as a vector
df = fromJSON(game)

eval_vector = df$eval

clk_vector = df$clk

print(eval_vector)
print(clk_vector)
```
Before we go any further, let's define how we'll interpret the change in evaluations from move to move. 
First off: any change is more significant when the evaluation is closer to 0. 
i.e. an eval going from +0 to +3 is more significant than +3 to +6, which is much more significant than +12 to +15.
Secondly, a shift in eval is considered equal magnitude regardless of the direction, i.e. +3 to +6 will be considerd identical in magnitude to +6 to +3
I will interpret MN (Math in n) or # as +20 / -20 for simplicity.
This metric I will call 'Adjusted Eval Delta'; for this I will use a logistic function to calculate an adjusted evaluation and calculate the differences:
$$ \frac{20}{1+e^{-0.25x}}\ -10 $$
From here I will take the absolute value of the difference to evaluate the quality of moves.

```{r}
convert_mate = function(x) {
  # If it doesn't contain '#', it's a normal centipawn evaluation
  if (!grepl("#", x)) {
    return(as.numeric(x))
  }
  
  if (grepl("-m", x)) {
    return (-20)
  } else {
    return (20)
  }
}

# Apply to vector
adjusted_eval = as.numeric(sapply(eval_vector, convert_mate))
adjusted_eval = adjusted_eval[!is.na(adjusted_eval)]
print(adjusted_eval)


```

```{r}

eval_logi = function(n){
  return (20 * plogis(0.25 * n) - 10)
}

ggplot(data.frame(x=c(-20, 20)), aes(x=x)) + 
  stat_function(fun=eval_logi)

#Take the change in evaluation per move
reg_diffs = diff(adjusted_eval)
adj_diffs = diff(eval_logi(adjusted_eval))

plot(reg_diffs, type = "o", cex = 0.5, col = "blue", xlab = "move number", ylab = "eval difference" ,ylim = c(-10, 20))
lines(adj_diffs, type = "o", cex = 0.5,  col = "red")
lines(adjusted_eval[-1], type = "o", cex = 0.5,  col = "green")
legend("topleft", legend = c("Regular Diffs", "Adjusted Diffs", "Evaluation"), 
       col = c("blue", "red", "green"),  lty = 1, pch = 1)
```
Now with this metric, the change of evaluation per move can be analyzed more precisely in regards to the actual affect on the outcome of the game. 

```{r}
mean(adj_diffs)
median(adj_diffs)
max(adj_diffs)
var(adj_diffs)
```

With this tool, we now can take aggregate functions over the entire dataset. To streamline this process, we add columns to the stockfish table containing the important per-game information.

```{r echo=FALSE, warning=FALSE}

means = NULL
medians = NULL
maxes = NULL
vars = NULL

for(i in 1:nrow(stockfish)){
  game = stockfish[i,'moves']
  
  df = fromJSON(game)
  eval = df$eval 
  
  transformed = eval |> 
    sapply(convert_mate) |>
    as.numeric() |>
    na.omit() |>
    eval_logi() |>
    diff() |>
    abs()
    
  
  means[i] = mean(transformed)
  medians[i] = median(transformed)
  maxes [i] = max(transformed)
  vars[i] = var(transformed)
  
}

stockfish$means = means
stockfish$medians = medians
stockfish$maxes = maxes
stockfish$vars = vars

stockfish$ratingdelta = abs(stockfish$white_elo - stockfish$black_elo)
stockfish$ratingavg = (stockfish$white_elo + stockfish$black_elo) / 2

```

```{r}
rating_groups <- stockfish %>%
  mutate(rating_bins = cut(ratingavg, 
                          breaks = seq(500, 3200, by = 900), 
                          include.lowest = TRUE, 
                          right = FALSE))


ggplot(rating_groups, aes(x = ratingdelta, y = vars, color = rating_bins)) +
  geom_point(alpha = 0.4, size = 1) +
  # Create separate visualizations for each 500-elo interval
  facet_wrap(~ rating_bins) + 
  # Trend line
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.5) +
  labs(
    title = "Eval Variance vs. Rating Difference by Elo Bracket",
    x = "Rating Difference",
    y = "Eval Variance" ) +
  theme_minimal() +
  theme(legend.position = "none")


```

```{r}
ggplot(rating_groups, aes(x = ratingdelta, y = means, color = rating_bins)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~ rating_bins) + 
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.5) +
  labs(
    title = "Mean Eval Change vs. Rating Difference by Elo Bracket",
    x = "Rating Difference",
    y = "Mean Eval Change" ) +
  theme_minimal() +
  theme(legend.position = "none")
```


```{r}
ggplot(rating_groups, aes(x = ratingdelta, y = medians, color = rating_bins)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~ rating_bins) + 
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.5) +
  labs(
    title = "Median Eval Change vs. Rating Difference by Elo Bracket",
    x = "Rating Difference",
    y = "Median Eval Change" ) +
  theme_minimal() +
  theme(legend.position = "none")
```


```{r}
ggplot(rating_groups, aes(x = ratingdelta, y = maxes, color = rating_bins)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~ rating_bins) + 
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.5) +
  labs(
    title = "Max Eval Change vs. Rating Difference by Elo Bracket",
    x = "Rating Difference",
    y = "Max Eval Change" ) +
  theme_minimal() +
  theme(legend.position = "none")
```
```{r}
rating_groups |> count(rating_bins)
```





```{r}

for(i in 1:nrow(stockfish)){
  game = stockfish[i,'moves']
  
  df = fromJSON(game)
  eval = df$eval 
  clk = period_to_seconds(hms(df$clk)) 
  
  transformed = eval |> 
    sapply(convert_mate) |>
    as.numeric() |>
    na.omit() |>
    diff() |>
    eval_logi()
}
```

```{r echo=FALSE, warning=FALSE}

columns = c("eval_change", "timespent", "rating_average", "ratingdelta", "time_control")
time_accuracy = data.frame(matrix(ncol = length(columns), nrow = 0))
colnames(time_accuracy) = columns

for(i in 1:nrow(stockfish)){
  game = stockfish[i,'moves']
  
  df = fromJSON(game)
  eval = df$eval 
  clk = period_to_seconds(hms(df$clk)) 
  
  transformed = eval |> 
    sapply(convert_mate) |>
    as.numeric() |>
    na.omit() |>
    eval_logi() |>
    diff() |>
    abs()

  timespent = diff(clk, lag = 2) * -1 
  
  #add these games to time_accuracy table:
  
  length(timespent) <- length(transformed) - 1
  
  temp = data.frame(eval_change = transformed[-1],
                    timespent = timespent, 
                    rating_average = stockfish[i, 'ratingavg'],
                    ratingdelta = abs(stockfish[i, 'ratingdelta']),
                    time_control = stockfish[i, 'time_control'])
  time_accuracy = rbind(time_accuracy, temp)
}


```

```{r}
unique(time_accuracy$time_control)
```
To normalize time controls, we will follow the standard of:
10+ minutes of time is rapid
3-10 minutes of time is blitz
Lower than 3 minutes is buillet
And total time is calculated by starting time + 40 * increment

```{r}
#Normalize time increment for analysis

for(i in 1:nrow(time_accuracy)){
  s = time_accuracy[i, 'time_control']
  
  #Calculate time format
  if(grepl('+', s, fixed = TRUE)){
    parts = as.numeric(unlist(strsplit(s, "\\+")))
    a = as.numeric(parts[1])
    b = as.numeric(parts[2])
    time = a + 40 * b
      
    if(time > 600) time_accuracy[i, 'time_control'] = "Rapid"
  
    else if(time > 180) time_accuracy[i, 'time_control'] = "Blitz"
  
    else time_accuracy[i, 'time_control'] = "Bullet"
  }
  
}
```

```{r}
#clean the data: remove NA and negative values for time spent
time_accuracy = time_accuracy |> filter(timespent >= 0)

```

```{r}
ggplot(time_accuracy, aes(x = timespent, y = eval_change, color = rating_average)) +
  geom_point(alpha = 0.4, size = 1) +
  scale_color_viridis_c() + 
  geom_smooth(method = "lm", color = "black", se = FALSE, linewidth = 0.5) +
  theme_minimal() +
  labs(color = "Average Rating")
```

```{r fig.height=6, fig.width=15}
ggplot(time_accuracy, aes(x = timespent, y = eval_change)) +
  facet_wrap(~ time_control, ncol = 3) +
  geom_bin2d(bins = 70) +
  scale_fill_viridis_c(trans = "log10") + 
  labs(
    fill = "log10(count)",
    title = "Evaluation Change by Time Spent",
    x = "Time Spent",
    y = "Eval Change") +
  theme_minimal()
```


